---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE, echo=FALSE,message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%")
```

# spAbundance 

```{r, echo=FALSE, results='asis', message=FALSE, eval = FALSE}
library(badger)
cat(
	# badger::badge_cran_release("spAbundance", "green"),
	# badger::badge_cran_download("spAbundance", "grand-total", "blue"),
	badger::badge_codecov()
)
```

`spAbundance` fits single-species and multi-species spatial abundance models using Markov Chain Monte Carlo (MCMC). The package is in initial stages of development and testing of existing functionality is not complete so use of the package prior to it's first release on CRAN is at your own risk. `spAbundance` uses analogous syntax to its "sister package" [spOccupancy](https://www.jeffdoser.com/files/spoccupancy-web/) (Doser et al. 2022). 

## Installation

`spAbundance` is in the initial stages of active development. The functionality currently implemented is not yet stable, and should be used with caution. To download the development version of the package, you can use `devtools` as follows:  

```{r, eval = FALSE}
devtools::install_github("doserjef/spAbundance")
```

Note that because we implement the MCMC in C++, you will need a C++ compiler on your computer to install the package from GitHub, which may not work on certain machines. If you run into this problem and are interested in using the development version of the package, feel free to send me an email: doserjef@msu.edu.

## Functionality

Below is a running list of currently implemented functionality. This will be updated as new functionality is included

|`spAbundance` Function  | Description                                                     |
|---------------- |------------------------------------------------------------------------|
|`NMix()`           | Single-species N-mixture model                                       |
|`spNMix()`         | Single-species spatial N-mixture model                               |
|`abund()`          | Single-species abundance GLM without imperfect detection             |
|`spAbund()`        | Single-species spatial abundance GLM without imperfect detection     |
|`msAbund()`        | Multi-species abundance GLM without imperfect detection              |
|`ppcAbund()`       | Posterior predictive check using Bayesian p-values                   | 
|`waicAbund()`      | Calculate Widely Applicable Information Criterion (WAIC)             |
|`simNMix()`        | Simulate single-species repeated count data                          |
|`simAbund()`       | Simulate single-species count data without imperfect detection       |

All model fitting functions allow for Poisson and negative binomial distributions for the abundance portion of the model. 

## Functionality to be included prior to first release

+ Single-species and multi-species spatial and non-spatial N-mixture models with Poisson and Negative Binomial distributions for latent abundance.
+ Single-species and multi-species spatial and non-spatial abundance GLMS without imperfect detection with Poisson and Negative Binomial likelihoods.
+ Single-species and multi-species spatial and non-spatial distance sampling models with Poisson and Negative Binomial distributions for latent abundance.
+ Posterior predcitve checks for all model types.
+ Simulation functions for all model types.
+ WAIC and k-fold cross-validation for all model types.

## References

Doser, J. W., Finley, A. O., Kéry, M., and Zipkin, E. F. (2022a). spOccupancy: An R package for single-species, multi-species, and integrated spatial occupancy models. Methods in Ecology and Evolution. https://doi.org/10.1111/2041-210X.13897.

